---
title: "PA1_template"
author: "Christer Smith"
date: "November 12, 2015"
output: html_document
---
#Data
Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

**steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)

**date**: The date on which the measurement was taken in YYYY-MM-DD format

**interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

##Loading packages, loading and preprocessing the data
```{r, echo=TRUE, warning=FALSE}
library(data.table)
library(ggplot2) # we shall use ggplot2 for plotting figures
running <- read.csv('activity.csv', header = TRUE, sep = ",",
colClasses=c("numeric", "character", "numeric"))
```
Converting the **date** field to a Date Class and **interval** field to a factor classes helps us organize our data and reference it later.
```{r,echo=TRUE,warning=FALSE}
running$date <- as.Date(running$date, format = "%Y-%m-%d")
running$interval <- as.factor(running$interval)
```
Let's look at the data to see how our organization looks:
```{r,echo=TRUE,warning=FALSE}
str(running)
```

##What is the mean total number of steps taken per day?

To get the steps taken per day, or daily steps, we looks at an aggregate of steps on each given day and sum them in each day.
```{r, echo=TRUE, warning=FALSE}
daily_steps<- aggregate(steps ~ date, running, sum)
colnames(daily_steps) <- c("date","steps")
head(daily_steps)
```
Next we look at the mean and median of the daily steps over the entire period. 
```{r,echo=TRUE,warning=FALSE}
mean_steps   <- mean(daily_steps$steps, na.rm=TRUE)
median_steps <- median(daily_steps$steps, na.rm=TRUE)
mean_steps 
median_steps
```
The mean and median are very close. If we show the mean and median on a historgraph, I bet it will be difficult to tell them apart. Let's graph them on the histograph below and see.
```{r,echo=TRUE,warning=FALSE}
p<-ggplot(daily_steps, aes(x = steps)) + 
       geom_histogram(fill = "orange", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Frequency of Days With Number of Steps") + theme_bw() 

pp<-p + geom_vline(xintercept = mean_steps, color="red") + geom_vline(xintercept = median_steps, color="blue")
pp
```

Yep, you can't tell that the median and mean are different visually, though according to our table above, the mean > median, suggesting positive (left) skew. 

##What is the average daily activity pattern?

To make a time-series plot of the 5-minute intervals, we should aggregate and find the mean of the 5-minute intervals and save them to a data frame.
```{r, echo=TRUE,warning=FALSE}
interval_steps <- aggregate(running$steps, by=list(interval = running$interval),FUN=mean, na.rm=TRUE)
#converting to integers to make plotting easier
interval_steps$interval <- 
        as.integer(levels(interval_steps$interval)[interval_steps$interval])
colnames(interval_steps) <- c("interval", "steps")
```
Time Series Plot of the average number of steps taken (averaged across all days) vs the five minute intervals.
```{r,echo=TRUE,warning=FALSE}
ggplot(interval_steps, aes(x=interval, y=steps)) +   
        geom_line(color="orange", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps Per 5-Minute Interval") +theme_bw()
```

We see above there is a maximum around the 800 interval. Let's check:
```{r, echo=TRUE,warning=FALSE}
max_interval <- interval_steps[which.max(  
        interval_steps$steps),]
max_interval
interval_steps$interval
```
The maximum is 206 steps at the 835th interval. From the chart above on the interval_steps$interval function, we see that the 835th interval is equivalent to 8 hours and 35 minutes, or following clock time, 8:35am.

##Inputing Missing Values:
**Total Number of Missing Values:**
```{r}
sum(is.na(running$steps))
```
2304 missing values out of a total of 17568 values is a 13.1% loss. We should compensate for the missing values to increasing the accuracy.

**Strategy for filling in all of the missing values in the dataset**

We replace the missing values with the mean value of at the same interval across all days. 
We create a function na-noNA(data,pervalue) where the data arguement is the running data frame and pervalue arguement is the inteval-steps data frame.
```{r, echo=TRUE,warning=FALSE}
na_noNA <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        noNA_steps <- data$steps
        noNA_steps[na_index] <- na_replace
        noNA_steps
}

running_noNA <- data.frame(  
        steps = na_noNA(running, interval_steps),  
        date = running$date,  
        interval = running$interval)
str(running_noNA)
```
This time, we see that the NA's are replaced by numbers, but we still have the same number of observations of 3 variables. Now, we should just check to make sure we didn't miss any of the NA's:
```{r,echo=TRUE,warning=FALSE}
sum(is.na(running_noNA$steps))
```
We did not.

**A histogram of the total numbers of steps taken each day
Below is a histogram with our new data set:
```{r,echo=TRUE,warning=FALSE}
noNA_daily_steps <- aggregate(steps ~ date, running_noNA, sum)
colnames(noNA_daily_steps) <- c("date","steps")
noNA_mean_steps   <- mean(noNA_daily_steps$steps, na.rm=TRUE)
noNA_median_steps <- median(noNA_daily_steps$steps, na.rm=TRUE)

##plotting the histogram
q<-ggplot(noNA_daily_steps, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Frequency of Days with Number of Steps") + theme_bw() 
qq<-q+geom_vline(xintercept = noNA_mean_steps, color="red") + geom_vline(xintercept = noNA_median_steps, color="orange")
```
It looks a little different. Let's compare the two side by side to see what looks different:
```{r,echo=TRUE,warning=FALSE}
require(gridExtra)
grid.arrange(pp,qq,ncol=2)
```
It looks more normally distributed. The maximum is still highly obvious.

**Do these values differe from the estimates of the first part of the assignment?**
```{r,echo=TRUE,warning=FALSE}
df<-data.frame(c(mean_steps,median_steps),c(noNA_mean_steps,noNA_median_steps))
names(df)[1]<-"Before"
names(df)[2]<-"After"
rownames(df)[1]<-"Mean"
rownames(df)[2]<-"Median"
df
```
Yes, they do. The median now matches the mean. This would also explain why the historgram looks more even. 

**What is the impact of inputting missing data on the estimate of the total daily number of steps?**

The mean has not changed, but the median moved to match the mean. This suggests the mean has become a better approximator of the average of our data, and skew is eliminated. The maximum frequency of days with numbers of steps has increased. This suggests more of the NA data occurred during peak times. This could suggest problems with the device measurements during higher intensity periods.

**Are there differences in activity patterns between weekdays and weekends?**

We will use our new data set with fill-in missing values.
First, we will add a column to the table that indicates the day of the week.
```{r, echo=TRUE,warning=FALSE}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}
```
Then subset the table into two parts: weekends and weekdays.
Tabulate weekend-steps and weekday-steps (in same form as intefvalsteps earlier, but with a column designating for weekends and weekdays)
```{r,echo=TRUE,warning=FALSE}
data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(running_noNA)
```
And plot the two graphs next to each other. We can use the facet-wrap function because both charts draw from the same data set.
```{r,echo=TRUE,warning=FALSE}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="violet") + 
        facet_wrap(~ dayofweek, nrow=1, ncol=2) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```

We see above that weekdays have a higher maximum peak than weekends, but weekends have more consistently high levels throughout the days. If our subject is rushing to work in the morning, like many people do, and works in an office most of the day, then this would account for our weekday data. And on the weekends, more consistent effort on other things besides working in a office, would show as we expect. We also see that the number of steps drop off later in the evening on weekends than weekdays, which is to be expected.